Interfacing logic programming and Coq

### 2020-10-27 papers/types2020/cultures.tex
Finished a revision of the first two sections.

In proof theory, one learns quickly that things always come in pairs:
negative/positive, left/right, bottom-up/top-down,
premises/conclusion, introduction/elimination,
% determinism/nondeterminism,
% invertible/noninvertible, 
etc.
%
When we examine the larger setting of this project of linking a logic
programming engine with Coq and its kernel, we find a large number of
pairing that are valuable to explicitly describe.

@@

In many ways, (structural) proof theory is more elementary and
low-level than most approaches to type theory.  For example, type
theories usually answer the question ``What is a proof?'' with the
response ``a dependently typed $\lambda$-term''.  That is, when
describing a type theory, one usually decides that a proof is a
certain kind of term within the system.  In contrast, the proof theory
treats logical propositions and proofs as separate.  For example,
proof theory does not assume that there are terms within the logic
that describe proofs.

\begin{metanote}
  This paragraph seem to identify proof and type theory with sequents and nat ded and while
  I agree with the gist (although people such as Frank would argue that sequents are a way of organizing proof search for nat ded), we're not gonna make a lot of friends at types and more importantly is a big hammer to nail the point that logic programming is helpful with Coq --am 
\end{metanote}

Gentzen's discovery that the key to treating classical and
intuitionistic logics in the same proof system was the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This discovery lead him away from natural
deduction to the invention of multiple-conclusion sequent calculus, a
very important kind of proof that has not been given a satisfactory
representation in type theory.  This same innovation of Gentzen also
opened the way to the invention of another key proof theoretic
discovery, that of linear logic \cite{girard87tcs}.
Similarly, sequent calculus provides an elegant presentation for
linear logic while most treatments of linear logic in type theory
usually only capture linearity in a partial fashion.

Proof theory also has rather immediate and natural treatments for
co-induction (as well as induction: see
\cite{baelde12tocl,momigliano12jal}) whereas the treatment of
coinduction within type theory remains more challenging and is well
developed only for various syntactic conditions (see, for example,
\cite{bertot08entcs}).
\begin{metanote}
  The criticism applies to Coq, less to type theory, see co-patterns --am
\end{metanote}

@@

Gentzen-style proofs are used to model computation in at least two
different ways.  The functional programming paradigm, following the
Curry-Howard correspondence, models computation abstractly as the
$\beta$-reduction of natural deduction proof: that is, computation is
modeled using \emph{proof normalization}.  On the other hand, the
logic programming paradigm, following the notion of goal-directed
proof search, models computation abstractly as a regimented search of
cut-free sequent calculus proofs: that is, computation is modeled
using \emph{proof search}.

Proof search has proved to have some richness that is hard to capture
in the proof normalization setting.  In particular, Gentzen's
invention of \emph{eigenvariables} are a kind of a proof-level
binder.  In the proof normalization setting, such eigenvariables
are instantiated during proof normalization.  In contrast, during the
search for cut-free proofs, eigenvariables are part of the syntax of
terms and formulas.  As a result, they can be used in the treatment of
bindings in data structures more generally.  Hence, the logic
programming paradigm provides a completely natural, elegant, and
powerful method, called $\lambda$-tree syntax, for treating bindings
within syntax.

In this paper, we shall use two different proof systems.  The first
logic is first-order intuitionistic logic.  Here, predicates
are not inductively defined and quantification will be limited to
range over first-order term.  Actually, we shall examine only the Horn
clause fragment of that logic.
%
The second logic will allow for inductively defined predicates using a
fixed point notation.   Given again our restriction to essentially
Horn clause specifications, it is possible to view intuitionistic
reasoning with such specification as if they are actually
specification in $\mu$MALL, an extension of multiplicative additive
linear logic with the least and greatest fixed point operators
\cite{baelde12tocl,baelde07lpar,heath19jar}.
%
The difference between these two logics and associate proof theories
is reminiscent to the difference between the type \lsti{Prop}
(corresponding to undefined propositions) and Sets (corresponding to
inductively defined propositions).
\begin{metanote}
  Still not onboard with this --am
\end{metanote}

@@

Given that \lP and Coq are both examples of mixing the
$\lambda$-calculus with logic, it is important to understand their
differences.  The confusion around the term \emph{higher-order
abstract syntax} (HOAS), for example, helps to illustrate their
differences.  In the functional programming setting, including the Coq
system, the HOAS approach leads to encoding binding structures within
terms using functions.  Such encodings have results in surprisingly
complicated encodings, allowing, for example, for \emph{exotic terms}
\cite{despeyroux95tlca} and to structures on which induction is not
possible \cite{roeckl01fossacs}.  In the logic programming setting,
particularly \lP, HOAS is well supported in both the availability of
binders allowed with terms ($\lambda$-bindings), in formulas
(quantifiers), and in proofs (eigenvariables).  For this reason, the
term $\lambda$-tree syntax was introduced to name this particular take
on HOAS \cite{miller19jar}.  The Abella proof assistant
\cite{baelde14jfr} was designed, in part, to allow for inductive and
co-inductive proofs for specifications using the $\lambda$-tree syntax
approach.

Another difference between \lP and functional programming can be
illustrated by considering how they are used in the specification of
tactics that are popular in modern proof assistants.  In fact, the
origin story for the ML functional programming language was that ML
was the meta-language for implementing the LCF suite of tactics and
tacticals \cite{gordon79}.  In order to implement tactics, ML adopted
not only novel features such as polymorphically typed higher-order
functional programming but also the non-functional mechanisms of
failure and exception handling.  While \lP is also based on
polymorphically typed higher-order relational programming, it also
comes with a completely declarative version of failure and
backtracking search.  Combining those features along with its support
of unification (even in the presence of term-level bindings), \lP
provides a rather different approach to the specification of tactics
\cite{felty93jar}.

While \lP is a typed language, it only uses simple types and these are
more often used to indicate syntactic categories.  Any dependency
information needs to be captured using predicates and quantifiers.

>>>>

When studying structural proof theory, one learns quickly that many
concepts come in pairs: negative/positive, left/right,
bottom-up/top-down, premises/conclusion, introduction/elimination,
etc.  When we examine the larger setting of this project of linking a
logic programming engine with Coq and its kernel, we find a large
number of new pairing that is valuable to explicitly discuss.

@@

In many ways, proof theory is more elementary and low-level than most
approaches to type theory.  For example, type theories usually answer
the question ``What is a proof?'' with the response ``a dependently
typed $\lambda$-term''.  That is, when describing a type theory, one
usually decides that a proof is a certain kind of $\lambda$-term
within the system.  In contrast, the proof theory treats logical
propositions and proofs as separate.  For example, proof theory does
not assume that there are terms within the logic that describe proofs.

%% \begin{metanote}
%%   This paragraph seem to identify proof and type theory with sequents
%%   and nat ded and while I agree with the gist (although people such as
%%   Frank would argue that sequents are a way of organizing proof search
%%   for nat ded), we're not gonna make a lot of friends at types and
%%   more importantly is a big hammer to nail the point that logic
%%   programming is helpful with Coq --am
%% \end{metanote}

%% DM: I've tried to address your concerns by adding an additional
%% paragraph and references (below).

Gentzen's discovery that the key to treating classical and
intuitionistic logics in the same proof system was the invention of
the weakening and contraction structural rules on the right-side of
sequents~\cite{gentzen35}.  This invention lead him away from natural
deduction to multiple-conclusion sequent calculus, a very important
kind of proof that has not been given a satisfactory representation in
type theory.  This same innovation of Gentzen also opened the way to
the invention of another key proof-theoretic discovery, that of linear
logic \cite{girard87tcs}.  Similarly, sequent calculus provides an
elegant presentation of linear logic while most treatments of linear
logic in type theory usually only capture linearity in a partial
fashion.

As is often observed, however, sequent calculus is too low-level to be
used explicitly as capturing the ``essence of a proof''.  Recent
developments in focused proof systems \cite{andreoli92jlc,liang09tcs},
makes it possible to extract from sequent calculus not only natural
deduction proofs \cite{pimentel16lsfa}, but also proof nets
\cite{chaudhuri08tcs}, and Herbrand-style expansion trees
\cite{chaudhuri16jlc}.  The role of focused proof systems to
characterize classes of proof structures is described in the next
section and used in our two example applications of ELPI with Coq.

%% DM I no longer think this comment is useful for this paper, so I
%% delete it instead of trying to fix it.

%% Proof theory also has rather immediate and natural treatments for
%% co-induction (as well as induction: see
%% \cite{baelde12tocl,momigliano12jal}) whereas the treatment of
%% coinduction within type theory remains more challenging and is well
%% developed only for various syntactic conditions (see, for example,
%% \cite{bertot08entcs}).
%% \begin{metanote}
%%   The criticism applies to Coq, less to type theory, see co-patterns --am
%% \end{metanote}

@@

Gentzen-style proofs are used to model computation in at least two
different ways.  The functional programming paradigm, following the
Curry-Howard correspondence, models computation abstractly as 
$\beta$-reduction of natural deduction proof: that is, computation is
modeled as \emph{proof normalization}.  On the other hand, the
logic programming paradigm, following the notion of goal-directed
proof search, models computation abstractly as a regimented search of
cut-free sequent calculus proofs: that is, computation is modeled
as \emph{proof search}.

Proof search has features that are hard to capture in the proof
normalization setting.  In particular, Gentzen's \emph{eigenvariables}
are a kind of a proof-level binder.  In the proof normalization
setting, such eigenvariables are instantiated during proof
normalization.  However, during the search for cut-free proofs,
eigenvariables do not vary, and they are part of the syntax of terms
and formulas.  As a result, they can be used in the treatment of
bindings in data structures more generally.  Such eigenvariables can
be used by the logic programming paradigm to provide a natural,
elegant, and powerful computing method with bindings within syntax.

It is worth noting that the role of the cut rule and of
cut-elimination is different in these two approaches to computing.
In the proof normalization paradigm, the cut rule can be used to model
$\beta$ reduction, especially via the explicit substitution approach
\cite{lengrand06phd}. In the proof search paradigm, since computing involves
the search for cut-free proofs, the cut rule plays no part in the
performance of computation.  However, cut and cut-elimination can be
used to reason about computation: for example, cut-elimination can be
used to prove ``substitution lemmas for free'' \cite{gacek12jar}.

%% The following paragraph seems out-of-place so I delete it now.

%% In this paper, we shall use two different proof systems.  The first
%% logic is first-order intuitionistic logic.  Here, predicates
%% are not inductively defined and quantification will be limited to
%% range over first-order term.  Actually, we shall examine only the Horn
%% clause fragment of that logic.
%% %
%% The second logic will allow for inductively defined predicates using a
%% fixed point notation.   Given again our restriction to essentially
%% Horn clause specifications, it is possible to view intuitionistic
%% reasoning with such specification as if they are actually
%% specification in $\mu$MALL, an extension of multiplicative additive
%% linear logic with the least and greatest fixed point operators
%% \cite{baelde12tocl,baelde07lpar,heath19jar}.

@@

Given that \lP and Coq both result from a mixing of the
$\lambda$-calculus with logic, it is important to understand their
differences.  The confusion around the term \emph{higher-order
abstract syntax} (HOAS), for example, helps to illustrate their
differences.  In the functional programming setting, including the Coq
system, the HOAS approach leads to encoding binding structures within
terms using functions.  Such encodings are surprisingly complicated
since they allow for \emph{exotic terms} \cite{despeyroux95tlca} and
for structures on which induction is not possible
\cite{roeckl01fossacs}.  In the logic programming setting,
particularly \lP, HOAS is well supported since bindings are allowed
with terms ($\lambda$-bindings), in formulas (quantifiers), and in
proofs (eigenvariables).  For this reason, the term
\emph{$\lambda$-tree syntax} was introduced to name this particular
take on HOAS \cite{miller19jar}.  The Abella proof assistant
\cite{baelde14jfr} was designed, in part, to provide inductive and
co-inductive inference involving specifications using the
$\lambda$-tree syntax approach.

Another difference between \lP and functional programming can be
illustrated by considering how they are used in the specification of
tactics.  In fact, the origin story for the ML functional programming
language was that it was the meta-language for implementing the LCF
suite of tactics and tacticals \cite{gordon79}.  To implement tactics,
ML adopted not only novel features such as polymorphically typed
higher-order functional programming but also the non-functional
mechanisms of failure and exception handling.  While \lP is also based
on polymorphically typed higher-order relational programming, it also
comes with a completely declarative version of failure and
backtracking search.  Combining those features along with its support
of unification (even in the presence of term-level bindings), \lP
provides a rather different approach to the specification of tactics
\cite{felty93jar}.

While \lP is a typed language, it only uses simple, polymorphic types,
and these are more often used to indicate syntactic categories.  Any
dependency information needs to be captured using predicates and
quantifiers.

### 2020-10-27 papers/types2020/intro.tex
Finished a revision of the first two sections.

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol}.
More recently, Tassi has made ELPI available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions.  First,
ELPI adds a notion of constraints, which makes it more expressive than,
say, the Teyjus implementation of \lP \cite{nadathur99cade}.  Second,
the plugin version of ELPI comes equip with an API for accessing the
Coq environment and a quotation and anti-quotation syntax allowing for
Coq expressions to be mixed with \lP program elements.

ELPI provides a means for treating bindings and unification (of
structures containing bindings) that are convenient for various
meta-programming tasks in logic programming all things useful for
tactic programming in Coq (proof refinement).  For examples, ELPI
programmers are spared from dealing with low-level aspects of
representing bindings (e.g., De Bruijn indexes) while still having
clean and effective ways to manipulate binding structures.  In short,
ELPI appears to be a useful meta-language for Coq
\cite{coen19mscs,tassi18coqpl,tassi19itp}.  
Type inference is a well-known example of a meta-programming task for
which logic programming often provides immediate and elegant
implementations via the sound implementation of logic.  % In this paper,
% we present two additional project in which an implementation of proof
% search provides direct, elegant, and compact implementations of
% meta-programming tasks.
In this paper, we outline
two applications of the Coq plugin for ELPI that support this claim.
With these examples, we shall illustrate that ELPI is a useful
meta-language not just because for its treatment of bindings and
syntactic structures in general but also because it is actually based
on a sound implementation of a higher-order intuitionistic logic.
Before we present these two examples, we
first discuss the different world views that the integration of ELPI
and Coq forces us to confront.

>>>>

Recently, C. Sacerdoti Coen and E. Tassi developed the ELPI
implementation \cite{dunchev15lpar} of \lP \cite{miller12proghol} and
more recently, Tassi has made ELPI available as a plugin to the Coq
proof assistant \cite{tassi18coqpl}.  This implementation of \lP
extends earlier implementations in primarily two directions.  First,
ELPI adds a notion of constraints, which makes it more expressive than
the Teyjus implementation \cite{nadathur99cade} of \lP. Second, the
plugin version of ELPI comes equipped with an API for accessing the
Coq environment and a quotation and anti-quotation syntax for mixing
for Coq expressions with \lP program elements.

The logic programming interpreter that underlies ELPI provides several
convenient features for the kind of meta-programming tasks that can
arise within modern proof assistants.  For example, \lP provides a
declarative and direct treatment of abstract syntax that contains
bindings: that treatment includes capture-avoiding substitution,
unification, and recursive programming.  The ELPI spares the
programmer from dealing with low-level aspects of bindings
representation (e.g., De Bruijn indexes) while still having clean and
effective ways to manipulate binding structures.

Since relations (not functions) play a central role in \lP, ELPI is
capable to providing direct support for the many relations that arise
in proof assistants.  Such relations include typing (e.g.,
$M\colon\sigma$), evaluation (e.g., natural semantics specifications
\cite{kahn87stacs,hannan93jfp}), and interaction (e.g., structured
operational semantics \cite{plotkin81,milner89book}).  Felty has also
made the point that LCF-style tactics and tacticals can be given an
elegant and natural specification using the higher-order relational
specifications provided by \lP \cite{felty93jar}.  Some recent
implementations built using ELPI
\cite{coen19mscs,tassi18coqpl,tassi19itp} support the usefulness of
higher-order logic programming as a meta-programming language useful
for Coq.

In this paper, we present two applications of ELPI within Coq. With
these examples, we shall illustrate that ELPI is useful not only
because it has some useful meta-programming features (e.g., a builtin
treatment of bindings in syntax) but also because it soundly
implements a higher-order intuitionistic logic.  The latter aspect of
ELPI makes it possible to encode the proofs and proof theory of
specific subsets of the Coq logic.

Before we can present these examples, we first highlight the rather
striking differences in notions of computing and reasoning that arise
on each side of the ELPI-Coq API.   We will also present a quick
summary of the key proof theory concepts that are used by our example
applications.
